---
title: "SMOTE"
author: "Vivek P."
output: html_document
---

# Using SMOTE to tackle Imbalance

We got a very weak specificity in the previous model. Before getting the ensemble methods out, we tried to get the most out of the single trees. In particular, we tried to apply a technique for dealing with the imbalance.

In the literature, many methods have been devised to address imbalanced data \cite{unbalanced}. These can be grouped into 3 main categories [Chapter 5, 2]:
\begin{enumerate}
    \item \textbf{Sampling methods:} The training set is modified to produce a more balanced distribution that allows classifiers to perform in a similar manner to standard classification. These methods are sometimes called data-level modifications \cite{fernandez2018learning}.
    \item \textbf{Algorithm-level modifications:} The classification algorithm is modified to be more attuned to class imbalance.
    \item \textbf{Cost-sensitive learning:} This incorporates data-level and algorithm-level modifications by considering variable misclassification costs.
\end{enumerate}

We will not have time to look into all of these, so we will only look at sampling methods. We refer the reader to the book cited for more details.

## SMOTE Details

We discuss one sampling method called the \textit{Synthetic Minority Oversampling Technique} (SMOTE).  It is based on the creation of synthetic data by interpolating between minority samples using their nearest neighbours (see the KNN classifier for how we define neighbours). Note that we encode each feature $\boldsymbol{x}$ so that each $x_i \in \mathbb{R}$ so that interpolation is possible. The formal process works as follows, and is adapted from \cite[Chapter 5.4, 1]. First, an integer value $N$, the oversampling factor, is specified. Default behaviour in programs will usually choose this to balance the class distribution. Then, an iterative process is carried out. First, a minority class point $\boldsymbol{x}^*$ is selected at random from the training set. Next, its $K$ nearest neighbours are obtained. Finally, $N$ of these $K$ instances are randomly chosen (with repetitions allowed). Then, $N$ synthetic examples are created by random linear interpolation between $\boldsymbol{x}^*$ and each of the chosen nearest neighbours.

# Loading Data

We load the data using the code below. 

**NOTE**: Reusing the code from the script led to issues and `R` couldn't find the test and train data. I copied them into my personal directory to overcome the issue. You may need to do this too.

```{r}
# Load necessary packages
library(rpart)
library(caret)  # For confusion matrix and precision/recall calculations
library(pROC)   # For ROC curve
library(ggplot2) # For plotting
library(dplyr)   # For data manipulation

# Load the training and test datasets
X_train <- read.csv("X_train_smote.csv", row.names = 1)  # Load the SMOTE transformed training features
y_train <- read.csv("y_train_smote.csv", row.names = 1)  # Load the SMOTE transformed training labels
X_test <- read.csv("X_test.csv", row.names = 1)    # Load the test features
y_test <- read.csv("y_test.csv", row.names = 1)    # Load the test labels

# Combine X_train and y_train into one data frame for rpart
train_data <- cbind(X_train, y_train)

# Drop the "education" column from the training data if it exists
train_data <- train_data[, !names(train_data) %in% "education"]

# Fit the classification tree using rpart
fit <- rpart(income ~ ., data = train_data, method = "class", control = rpart.control(cp = 1e-6))

# Get the cost-complexity pruning table and identify the best cp based on minimum xerror
cptable <- fit$cptable
best_cp <- cptable[which.min(cptable[,"xerror"]), "CP"]
cat("Best CP:", best_cp, "\n")

# Prune the tree using the best cp
pruned_tree <- prune(fit, cp = best_cp)
```

```{r}
# Load necessary libraries
library(ggplot2)    # For plotting
library(pROC)       # For ROC curve calculations
library(dplyr)      # For data manipulation

# Get predicted probabilities for the positive class
pred_probs <- predict(pruned_tree, X_test, type = "prob")[, "<=50K"]  # Adjust based on your positive class

# Compute ROC curve for the model with predictions
roc_curve_model <- roc(y_test[, 1], pred_probs, levels = c("<=50K", ">50K"))

# Calculate AUC for the model
auc_value_model <- auc(roc_curve_model)

# Print AUC value for the model
cat("AUC for Model:", auc_value_model, "\n")

# Create a data frame for the model's ROC data
roc_data_model <- data.frame(
  FPR = 1 - roc_curve_model$specificities,  # False Positive Rate
  TPR = roc_curve_model$sensitivities         # True Positive Rate
)

# Load the saved ROC data
roc_data_saved <- read.csv("roc_data_no_smote.csv")

# Ensure the saved data has columns for FPR and TPR
if (!all(c("FPR", "TPR") %in% colnames(roc_data_saved))) {
  stop("The saved ROC data must contain 'FPR' and 'TPR' columns.")
}

# Check and convert the columns to numeric (was getting errors)
roc_data_saved$FPR <- as.numeric(roc_data_saved$FPR)
roc_data_saved$TPR <- as.numeric(roc_data_saved$TPR)
roc_data_model$FPR <- as.numeric(roc_data_model$FPR)
roc_data_model$TPR <- as.numeric(roc_data_model$TPR)

# Create the plot for the saved ROC curve
plot(roc_data_saved$FPR, roc_data_saved$TPR, type = "l", col = "red", lwd = 2, 
     xlim = c(0, 1), ylim = c(0, 1), 
     main = "ROC Curves Comparison", 
     xlab = "False Positive Rate", 
     ylab = "True Positive Rate")

# Add the model ROC curve to the same plot
lines(roc_data_model$FPR, roc_data_model$TPR, col = "blue", lwd = 2)

# Add a diagonal line for random guessing
abline(0, 1, lty = 2, col = "grey")

# Add a legend to the plot
legend("bottomright", legend = c("ROC Curve (No Smote)", "ROC Curve (SMOTE)"),
       col = c("red", "blue"), lwd = 2)

```

This didn't make a notable difference. Reaching this point, we wanted to try more methods to deal with missing data, but this time using the SMOTE data only. We leave this to the next section.

[1] Fernández, Alberto, et al. Learning from imbalanced data sets. Vol. 10. No. 2018. Cham: Springer, 2018.